---
title: Mapping Healthy Ride Data
author: Conor Tompkins
date: '2017-11-04'
slug: mapping-healthy-ride-data
categories:
  - Pittsburgh
  - R
tags:
  - Healthy Ride
  - Pittsburgh
  - R Markdown
  - WPRDC
runtime: shiny
---

This post explores the Healthy Ride dataset using network analysis. 

This is my third post about the Healthy Ride bike service in Pittsburgh. You can find the [first post](https://ctompkins.netlify.com/2017/10/18/exploring-healthy-ride-pittsburgh-data/) and [second post](http://ctompkins.netlify.com/2017/11/05/healthy-ride-network-analysis/) on my blog.

Load the R packages we will be using:
```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggmap)
library(lubridate)
library(viridis)
library(stringr)
library(gghighlight)
```

Load the data from the WPRDC (hosted on [my GitHub page](https://github.com/conorotompkins)):
```{r load data, message=FALSE, warning=FALSE}
data <- read_csv("https://raw.githubusercontent.com/conorotompkins/healthy_ride/master/data/data.csv")
```

Again, we need to format the data and the column names to make them more useful for analysis. Since this is a repeat of the script from the last post, I will just do it all in one go:
```{r munge_data, message=FALSE, warning=FALSE}
colnames(data) <- tolower(colnames(data))
colnames(data) <- gsub(" ", "_", colnames(data))

data_long <- data %>% 
  rename(start_date_time = starttime,
         stop_date_time = stoptime) %>% 
  gather(date_time_type, date_time, c(start_date_time, stop_date_time)) %>% 
  select(date_time_type, date_time, everything()) %>% 
  mutate(date_time_2 = date_time) %>% 
  separate(date_time, " ", into = c("date", "time")) %>% 
  mutate(id = row_number(),
         date = mdy(date),
         year = year(date),
         month = month(date, label = TRUE),
         week = week(date),
         time = hm(time),
         hour = hour(time),
         wday = wday(date, label = TRUE),
         is_weekday = ifelse(wday %in% c("Mon", "Tues", "Wed", "Thurs", "Fri"), "weekday", "weekend"),
         yday = yday(date),
         mday = mday(date)) %>% 
  mutate(trip_duration = (tripduration / 60) / 60) %>% 
  gather(station_id_type, station_id, c(from_station_id, to_station_id)) %>% 
  gather(station_name_type, station_name, c(from_station_name, to_station_name)) %>% 
  select(date_time_type, 
         is_weekday, 
         date, 
         year,
         month,
         time, 
         hour,
         wday,
         yday,
         mday,
         date_time_2, 
         station_id_type, 
         station_id, 
         station_name_type,
         station_name,
         everything())
```

Importantly, we will be excluding trips where the rider started and ended their trip at the same station from this analysis. The data lacks the granularity to analyze rider location beyond the points where they began and ended their trip.

```{r}
data_long <- data_long %>% 
  spread(station_name_type, station_name) %>% 
  filter(from_station_name != to_station_name) %>% 
  gather(station_name_type, station_name, c(from_station_name, to_station_name))
```

We also need to load the CSV with the longitude and latitude for the Healthy Ride stations
```{r load_station_location, message=FALSE, warning=FALSE}
data_station_locations <- read_csv("https://raw.githubusercontent.com/conorotompkins/healthy_ride/master/data/stations/station_locations.csv")
```

```{r munge_station_location_data, message=FALSE, warning=FALSE}
(df_station_totals <- data_long %>% 
  group_by(station_name) %>% 
  summarize(number_of_trips = n()) %>% 
  arrange(desc(number_of_trips), station_name) %>% 
  left_join(data_station_locations) %>% 
  select(station_name, number_of_trips, longitude, latitude))
```

Where are the Healthy Ride Stations?
```{r}
pgh_map <- get_map(location = "The Hill Pittsburgh, PA", zoom = 13)
pgh_map <- ggmap(pgh_map)

pgh_map +
  geom_point(data = df_station_totals, aes(longitude, latitude, size = number_of_trips),
             alpha = .75) +
  scale_size_continuous(range = c(.1, 5)) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```

Create a dataframe with station name, number of rides, and longitude and latitude.
```{r}
df_long <- data_long %>% 
  select(station_name, station_name_type) %>% 
  group_by(station_name, station_name_type) %>% 
  summarize(number_of_trips = n()) %>% 
  arrange(desc(number_of_trips)) %>% 
  left_join(data_station_locations) %>% 
  ungroup()
```

Do some stations function more as starting points or ending points for trips?
```{r}
pgh_map +
  geom_point(data = df_long, aes(longitude, latitude, size = number_of_trips, color = station_name_type),
             alpha = .75) +
  scale_size_continuous(range = c(.1, 5)) +
  facet_wrap(~station_name_type) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```

No differences are discernible in this view.

A scatter plot shows the differences more effectively:
```{r}
df_from_to <- df_long %>%
  spread(station_name_type, number_of_trips) %>% 
  rename(from_trips = from_station_name,
         to_trips = to_station_name) %>% 
  select(from_trips, to_trips) %>% 
  mutate(differential = abs(from_trips - to_trips))

#use the stations with large differentials as jumping off points in the network map
df_from_to %>% 
  gghighlight_point(aes(from_trips, to_trips), label_key = station_name, differential > 4000) +
  scale_x_continuous(limits = c(0, 45000)) +
  scale_y_continuous(limits = c(0, 45000)) +
  coord_equal() +
  geom_abline() +
  theme_bw()
```
The top 10 stations in terms of absolute difference between departures and arrivals:
```{r}
df_from_to %>% 
  ungroup() %>% 
  arrange(desc(differential)) %>% 
  top_n(10, differential)
```

Let's map the connections between stations by drawing lines between the stations.

First, widen the data:
```{r}
df_wide <- data_long %>%
  spread(station_name_type, station_name) %>% 
  select(from_station_name, to_station_name) %>% 
  left_join(data_station_locations, by = c("from_station_name" = "station_name")) %>%
  rename(from_latitude = latitude,
         from_longitude = longitude) %>% 
  left_join(data_station_locations, by = c("to_station_name" = "station_name")) %>% 
  rename(to_latitude = latitude,
         to_longitude = longitude) %>% 
  group_by(from_station_name, to_station_name, from_longitude, from_latitude, to_longitude, to_latitude) %>% 
  summarise(number_of_trips = n()) %>% 
  arrange(desc(number_of_trips)) %>% 
  mutate(from_station_type = ifelse(from_station_name == to_station_name,
                               "Same station", "Different station"))
```

Then, layer the data over the map:
```{r}
#to-from map
pgh_map +
  geom_segment(data = df_wide, aes(x = from_longitude, xend = to_longitude, 
                                   y = from_latitude, yend = to_latitude, 
                                   alpha = number_of_trips, size = number_of_trips)) +
  geom_point(data = df_wide, aes(from_longitude, from_latitude), shape = 21, size = 3, fill = "white") +
  geom_point(data = df_wide, aes(to_longitude, to_latitude), shape = 21, size = 3, fill = "white") +
  scale_alpha_continuous(range = c(.05, .3)) +
  scale_size_continuous(range = c(.05, 3)) +
  #facet_wrap(~from_station_name) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
#ggsave("images/ride_map.png")
```

We can also facet by the from_station_name variable to see where trips originating from certain stations end at. This plot shows the top 6 stations in terms of trips that began from that station:
```{r}
top_from_stations <- df_wide %>% 
  group_by(from_station_name) %>% 
  summarize(number_of_trips = sum(number_of_trips)) %>% 
  arrange(desc(number_of_trips)) %>% 
  top_n(6) %>% 
  select(from_station_name) %>% 
  unlist()
top_from_stations

df_wide_specific_station <- df_wide %>% 
  filter(from_station_name %in% top_from_stations)

pgh_map +
  geom_point(data = df_wide_specific_station, aes(from_longitude, from_latitude), 
             shape = 21, size = 3, fill = "white") +
  geom_point(data = df_wide_specific_station, aes(to_longitude, to_latitude), 
             shape = 21, size = 3, fill = "white") +
  geom_segment(data = df_wide_specific_station, aes(x = from_longitude, xend = to_longitude, 
                                   y = from_latitude, yend = to_latitude, 
                                   alpha = number_of_trips, size = number_of_trips),
               arrow = arrow(length = unit(0.03, "npc"))) +
  scale_alpha_continuous(range = c(.3, .5)) +
  scale_size_continuous(range = c(.05, 3)) +
  facet_wrap(~from_station_name,
             ncol = 3) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```


looking at top stations in terms of absolute difference
```{r}
top_diff_stations <- df_from_to %>% 
  arrange(desc(differential)) %>% 
  top_n(10) %>% 
  select(station_name) %>% 
  unlist()
  
top_diff_stations

df_wide_diff_station <- df_wide %>% 
  filter(from_station_name %in% top_diff_stations)

pgh_map +
  geom_point(data = df_wide_diff_station, aes(from_longitude, from_latitude), 
             shape = 21, size = 3, fill = "white") +
  geom_point(data = df_wide_diff_station, aes(to_longitude, to_latitude), 
             shape = 21, size = 3, fill = "white") +
  geom_segment(data = df_wide_diff_station, aes(x = from_longitude, xend = to_longitude, 
                                   y = from_latitude, yend = to_latitude, 
                                   alpha = number_of_trips, size = number_of_trips),
               arrow = arrow(length = unit(0.03, "npc"))) +
  scale_alpha_continuous(range = c(.3, .5)) +
  scale_size_continuous(range = c(.05, 3)) +
  facet_wrap(~from_station_name,
             ncol = 3) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```



```{r}
df_wide_day <- data_long %>% 
  spread(station_name_type, station_name) %>% 
  select(from_station_name, to_station_name, is_weekday) %>% 
  left_join(data_station_locations, by = c("from_station_name" = "station_name")) %>%
  rename(from_latitude = latitude,
         from_longitude = longitude) %>% 
  left_join(data_station_locations, by = c("to_station_name" = "station_name")) %>% 
  rename(to_latitude = latitude,
         to_longitude = longitude) %>% 
  group_by(is_weekday, from_station_name, to_station_name, from_longitude, from_latitude, to_longitude, to_latitude) %>% 
  summarise(number_of_trips = n()) %>% 
  arrange(desc(number_of_trips))

pgh_map +
  geom_segment(data = df_wide_day, aes(x = from_longitude, xend = to_longitude, 
                                   y = from_latitude, yend = to_latitude, 
                                   alpha = number_of_trips, size = number_of_trips)) +
  geom_point(data = df_wide_day, aes(from_longitude, from_latitude), shape = 21, size = 3, fill = "white") +
  geom_point(data = df_wide_day, aes(to_longitude, to_latitude), shape = 21, size = 3, fill = "white") +
  scale_alpha_continuous(range = c(.05, .3)) +
  scale_size_continuous(range = c(.05, 3)) +
  facet_wrap(~is_weekday) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```

```{r}
df_wide_tod <- data_long %>% 
  spread(station_name_type, station_name) %>% 
  select(from_station_name, to_station_name, hour) %>% 
  mutate(time_of_day = cut(hour, breaks = c(-Inf, 3, 6, 9, 12, 15, 18, 21, Inf), 
                           labels = c("0-3", "3-6", "6-9", "9-12", "12-15", "15-18", "18-21", "21-24"), 
                           ordered_result = TRUE)) %>% 
  left_join(data_station_locations, by = c("from_station_name" = "station_name")) %>%
  rename(from_latitude = latitude,
         from_longitude = longitude) %>% 
  left_join(data_station_locations, by = c("to_station_name" = "station_name")) %>% 
  rename(to_latitude = latitude,
         to_longitude = longitude) %>% 
  group_by(time_of_day, from_station_name, to_station_name, 
           from_longitude, from_latitude, 
           to_longitude, to_latitude) %>% 
  summarise(number_of_trips = n()) %>% 
  arrange(desc(number_of_trips))

pgh_map +
  geom_segment(data = df_wide_tod, aes(x = from_longitude, xend = to_longitude, 
                                   y = from_latitude, yend = to_latitude, 
                                   alpha = number_of_trips, size = number_of_trips)) +
  geom_point(data = df_wide_tod, aes(from_longitude, from_latitude), shape = 21, size = 2, fill = "white") +
  geom_point(data = df_wide_tod, aes(to_longitude, to_latitude), shape = 21, size = 2, fill = "white") +
  scale_alpha_continuous(range = c(.05, .3)) +
  scale_size_continuous(range = c(.05, 3)) +
  facet_wrap(~time_of_day) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
```