---
title: Healthy Ride Network Analysis
author: Conor Tompkins
date: '2017-10-23'
slug: healthy-ride-network-analysis
categories:
  - Pittsburgh
  - R
tags:
  - Pittsburgh
  - WPRDC
  - Healthy Ride
draft: true
runtime: shiny
---

```{r initial options, message=FALSE, warning=FALSE, include=FALSE}
options(scipen = 999)
```

This post is about exploring the Healthy Ride dataset using network analysis. 

This is my second post about the Healthy Ride bike service in Pittsburgh. You can find the first post [here](https://ctompkins.netlify.com/2017/10/18/exploring-healthy-ride-pittsburgh-data/), where I did some exploratory analysis of the data.

Load the packages we will be using:
```{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggraph)
library(igraph)
library(lubridate)
library(viridis)
library(stringr)
```


Set the theme to format the plots:
```{r set theme}
theme_set(theme_graph())
```

Load the data from the WPRDC (hosted on [my GitHub page](https://github.com/conorotompkins)):
```{r load data}
data <- read_csv("https://raw.githubusercontent.com/conorotompkins/healthy_ride/master/data/data.csv")
```


Again, we need to format the data and the column names to make them more useful for analysis. Since this is a repeat of the script from the last post, I will just do it all in one go:
```{r munge 1}
colnames(data) <- tolower(colnames(data))
colnames(data) <- gsub(" ", "_", colnames(data))

data_long <- data %>% 
  rename(start_date_time = starttime,
         stop_date_time = stoptime) %>% 
  gather(date_time_type, date_time, c(start_date_time, stop_date_time)) %>% 
  select(date_time_type, date_time, everything()) %>% 
  mutate(date_time_2 = date_time) %>% 
  separate(date_time, " ", into = c("date", "time")) %>% 
  mutate(id = row_number(),
         date = mdy(date),
         year = year(date),
         month = month(date, label = TRUE),
         week = week(date),
         time = hm(time),
         hour = hour(time),
         wday = wday(date, label = TRUE),
         is_weekday = ifelse(wday %in% c("Mon", "Tues", "Wed", "Thurs", "Fri"), "weekday", "weekend"),
         yday = yday(date),
         mday = mday(date)) %>% 
  mutate(trip_duration = (tripduration / 60) / 60) %>% 
  gather(station_id_type, station_id, c(from_station_id, to_station_id)) %>% 
  gather(station_name_type, station_name, c(from_station_name, to_station_name)) %>% 
  select(date_time_type, 
         is_weekday, 
         date, 
         year,
         month,
         time, 
         hour,
         wday,
         yday,
         mday,
         date_time_2, 
         station_id_type, 
         station_id, 
         station_name_type,
         station_name,
         everything())
```

Check that the data is OK:
```{r preview data}
data_long
```

To make the data more useful for network analysis, we need to make it a bit wider:
```{r make data wide}
(data_wide <- data_long %>% 
  spread(station_name_type, station_name))
```

Set the seed for the random number generator so the charts look the same every time they are run:
```{r set seed}
set.seed(12345)
```

To start, lets look at the entire network:
```{r message=FALSE, warning=FALSE}
(simple_network <- data_wide %>% 
  count(from_station_name, to_station_name, sort = TRUE))

simple_network <- simple_network %>% 
  graph_from_data_frame()

ggraph(simple_network) +
  geom_edge_fan(aes(edge_alpha = n,
                    edge_width = n)) +
  geom_edge_loop(aes(edge_alpha = n,
                     edge_width = n)) +
  geom_node_point() +
  scale_edge_alpha_continuous(range = c(.1, 1)) +
  scale_edge_width_continuous(range = c(.5, 1))
```

This is a bit of a hairball, and is frankly not useful.

Instead, lets look at how the top 400 combinations of departures and arrivals are networked:
```{r top 200, message=FALSE, warning=FALSE}
simple_network <- data_wide %>% 
  count(from_station_name, to_station_name, sort = TRUE) %>% 
  top_n(400)

simple_network <- simple_network %>% 
  graph_from_data_frame()

ggraph(simple_network) +
  geom_edge_fan(aes(edge_alpha = n,
                    edge_width = n)) +
  geom_edge_loop(aes(edge_alpha = n,
                     edge_width = n)) +
  geom_node_point() +
  scale_edge_alpha_continuous(range = c(.3, 1)) +
  scale_edge_width_continuous(range = c(.5, 1))
```

There appear to be two clusters of stations.

We can also do a more traditional heatmap of the departure and arrival stations:
```{r top 200 weekday, message=FALSE, warning=FALSE}
simple_network <- data_wide %>% 
  #filter(from_station_name %in% top_from_stations,
  #       to_station_name %in% top_to_stations) %>% 
  count(from_station_name, to_station_name, sort = TRUE) %>% 
  top_n(400) %>% 
  ungroup() %>% 
  complete(from_station_name, to_station_name) %>% 
  replace_na(list(n = 0)) %>% 
  mutate(from_station_name_abbr = str_c(str_sub(from_station_name, 1, 18), "..."),
         to_station_name_abbr = str_c(str_sub(to_station_name, 1, 18), "..."))
#add complete and replace_na

length(unique(simple_network$from_station_name) == unique(simple_network$from_station_name_abbr))
length(unique(simple_network$to_station_name)) == length(unique(simple_network$to_station_name_abbr))


x_axis <- simple_network %>% 
  group_by(from_station_name_abbr) %>% 
  summarize(n = sum(n)) %>% 
  arrange(desc(n)) %>%
  select(from_station_name_abbr) %>% 
  unlist()
  
y_axis <- simple_network %>% 
  group_by(to_station_name_abbr) %>% 
  summarize(n = sum(n)) %>% 
  arrange(desc(n)) %>%
  select(to_station_name_abbr) %>% 
  unlist()

simple_network %>% 
  mutate(from_station_name_abbr = factor(from_station_name_abbr, levels = x_axis),
         to_station_name_abbr = factor(to_station_name_abbr, levels = y_axis)) %>% 
  ggplot(aes(from_station_name_abbr, to_station_name_abbr, fill = n)) +
  geom_tile() +
  theme_bw() +
  coord_equal() +
  scale_fill_viridis() +
  #scale_x_discrete(breaks = unique(simple_network$from_station_name_abbr)) +
  guides(fill = guide_colorbar(title = "Number of Rides")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0))
```

It appears that many trips start and end at the same station.

We can investigate this phenomenon:
```{r}
data_wide %>% 
  select(from_station_name, to_station_name) %>% 
  mutate(is_same = ifelse(from_station_name == to_station_name, "same_location", "different_location")) %>%
  group_by(is_same) %>% 
  count() %>% 
  spread(is_same, n) %>% 
  gather(location_type, n) %>% 
  mutate(from_location = "from_location") %>% 
  select(from_location, location_type, n) %>% 
  mutate(n = round(n / sum(n), digits = 2))
```
27% of all trips end at the same station they began at.

Does the network look different on weekdays vs. weekends?
```{r top_200_weekday, message=FALSE, warning=FALSE}
simple_network <- data_wide %>% 
  count(from_station_name, to_station_name, is_weekday, sort = TRUE) %>% 
  top_n(400)

simple_network <- simple_network %>% 
  graph_from_data_frame()

ggraph(simple_network) +
  geom_edge_fan(aes(edge_alpha = n,
                    edge_width = n,
                    color = is_weekday)) +
  geom_edge_loop(aes(edge_alpha = n,
                     edge_width = n,
                     color = is_weekday)) +
  geom_node_point() +
  scale_edge_alpha_continuous(range = c(.3, 1)) +
  scale_edge_width_continuous(range = c(.5, 1)) +
  facet_edges(~is_weekday) +
  th_foreground(foreground = 'grey80', border = TRUE)
```

The network appears to be smaller and more compact on the weekends


```{r}
df_test <- data_wide %>% 
  select(from_station_name, to_station_name, is_weekday) %>% 
  mutate(is_same = ifelse(from_station_name == to_station_name, "same_location", "different_location")) %>%
  group_by(is_same, is_weekday) %>% 
  count() %>% 
  spread(is_same, n) %>% 
  gather(location_type, n, -is_weekday) %>% 
  mutate(from_location = "from_location") %>% 
  select(is_weekday, from_location, location_type, n) %>% 
  mutate(n = round(n / sum(n), digits = 2)) %>% 
  arrange(is_weekday)

df_test %>% 
  ggplot(aes(location_type, n, fill = is_weekday)) +
  geom_col(position = "dodge") +
  labs(y = "% of rides") +
  theme_bw()
```

A greater percentage of trips start and end at the same station on the weekends than on weekdays

```{r}
#add network chart split by weekday/weekend and time of day (early morning, midday, evening, night)
#cut hour column
unique(data_wide$hour)

simple_network <- data_wide %>% 
  mutate(time_of_day = cut(hour, breaks = c(0, 6, 12, 18, 23), labels = c("early morning", "Mid day", "Evening", "Night time"), ordered_result = TRUE)) %>% 
  count(from_station_name, to_station_name, is_weekday, time_of_day, sort = TRUE) %>% 
  filter(n > 100) %>% 
  mutate(time_of_day = as.character(time_of_day))



#sum(is.na(simple_network))

#str(simple_network)

simple_network <- simple_network %>% 
  graph_from_data_frame()
#str(simple_network)

ggraph(simple_network) +
  geom_edge_fan(aes(edge_alpha = n,
                    edge_width = n,
                    color = is_weekday)) +
  geom_edge_loop(aes(edge_alpha = n,
                     edge_width = n,
                     color = is_weekday)) +
  geom_node_point() +
  scale_edge_alpha_continuous(range = c(.3, 1)) +
  scale_edge_width_continuous(range = c(.5, 1)) +
  facet_graph(is_weekday ~ time_of_day)
#faceting not working
```