---
title: Exploring Healthy Ride Pittsburgh Data
author: Conor Tompkins
date: '2017-10-18'
slug: exploring-healthy-ride-pittsburgh-data
categories:
  - Pittsburgh
  - R
tags:
  - Pittsburgh
  - R Markdown
  - WPRDC
draft: true
runtime: shiny
---

Healthy Ride is a bike rental service in Pittsburgh. The [WPRDC](http://www.wprdc.org/) hosts [data about their service](https://data.wprdc.org/organization/about/healthy-ride). We can use the data to get an idea of how Pittsburghers use the service.

In my analysis, I use R, the tidyverse package, rmarkdown, and some other packages.

First, load the packages we will be using:
```{r load packages}
library(tidyverse)
library(lubridate)
library(viridis)
```

Set the theme to format the plots:
```{r set theme}
theme_set(theme_bw())
```

Load the data from the WPRDC (hosted on [my GitHub page](https://github.com/conorotompkins)):
```{r load data}
data <- read_csv("https://raw.githubusercontent.com/conorotompkins/healthy_ride/master/data/data.csv")
```

Format the data and the column names to make them more useful for analysis:
```{r format data}
colnames(data) <- tolower(colnames(data))
colnames(data) <- gsub(" ", "_", colnames(data))

data_long <- data %>% 
  mutate(start_date_time = starttime,
         stop_date_time = stoptime) %>%
  select(-c(starttime, stoptime)) %>% 
  gather(date_time_type, date_time, c(start_date_time, stop_date_time)) %>% 
  select(date_time_type, date_time, everything()) %>% 
  mutate(date_time_2 = date_time) %>% 
  separate(date_time, " ", into = c("date", "time")) %>% 
  mutate(id = row_number(),
         date = mdy(date),
         year = year(date),
         month = month(date, label = TRUE),
         week = week(date),
         time = hm(time),
         hour = hour(time),
         wday = wday(date, label = TRUE),
         is_weekday = ifelse(wday %in% c("Mon", "Tues", "Wed", "Thurs", "Fri"), "weekday", "weekend"),
         yday = yday(date),
         mday = mday(date)) %>% 
  mutate(trip_duration = (tripduration / 60) / 60) %>% 
  gather(station_id_type, station_id, c(from_station_id, to_station_id)) %>% 
  gather(station_name_type, station_name, c(from_station_name, to_station_name)) %>% 
  select(date_time_type, 
         is_weekday, 
         date, 
         year,
         month,
         time, 
         hour,
         wday,
         yday,
         mday,
         date_time_2, 
         station_id_type, 
         station_id, 
         station_name_type,
         station_name,
         everything())
```

Let's look at how the data look:L
```{r preview data}
data_long
```

How does the use of the service trended in the long term?
```{r long term density line plot}
data_long %>% 
  filter(date_time_type == "start_date_time") %>% 
  ggplot(aes(date)) +
  geom_freqpoly(stat = "density")
```

A cumulative look at the number of rides:
```{r long term cumulative line plot}
data_long %>% 
  filter(station_name_type == "from_station_name") %>% 
  arrange(date) %>% 
  count(date) %>% 
  mutate(cum_sum = cumsum(n)) %>% 
  ggplot(aes(date, cum_sum)) +
  geom_line()
```

A yearly look at the data shows that ridership has dropped off in 2017:
```{r yearly cumulative line plot}
data_long %>%
  select(year, yday) %>% 
  mutate(year = as.factor(year)) %>% 
  group_by(year, yday) %>% 
  count() %>% 
  ungroup() %>% 
  group_by(year) %>% 
  mutate(cum_sum = cumsum(n)) %>% 
  ggplot(aes(yday, cum_sum,
             color = year,
             group = year)) +
  geom_line()
```

How are the number of rides distributed throughout the day?
```{r hour plot}
data_long %>% 
  filter(date_time_type == "start_date_time") %>% 
  ggplot(aes(hour)) +
  geom_freqpoly(stat = "count")
```

Ridership peaks during evening rush-hour

The trend is different for weekdays vs. weekends:
```{r hour weekday vs weekend line plot}
data_long %>% 
  filter(date_time_type == "start_date_time") %>% 
  ggplot(aes(hour, color = is_weekday)) +
  geom_freqpoly(stat = "count")
```

Breaking it down by weekday, people appear to take rides during their lunch break on Fridays. Saturdays at around 1PM see peak usage.
```{r weekday hour line plot}
data_long %>%
  filter(date_time_type == "start_date_time") %>% 
  ggplot(aes(hour, color = wday)) +
  geom_freqpoly(stat = "count") +
  scale_x_continuous(breaks = c(0:23))
```

This tile chart shows the number of rides throughout the history of the data by year, month, day of the month, and hour:
```{r count tile plot}
data_long %>% 
  #need to add complete() and replace_na() to insert 0s
  group_by(year, month, mday, hour) %>% 
  summarize(n = n()) %>% 
  ggplot(aes(mday, hour, fill = n)) +
  geom_tile() +
  scale_y_reverse(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +
  facet_grid(year ~ month) +
  coord_equal() +
  scale_fill_viridis()
#need to make year strip text go horizontal
#maybe make this figure larger
```

This tile chart shows the sum of hours biked across the same date and time variables:
```{r trip duration tile plot}
data_long %>%
  #need to add complete() and replace_na() to insert 0s
  select(year, month, mday, hour, trip_duration) %>% 
  group_by(year, month, mday, hour) %>% 
  summarize(trip_duration_sum = sum(trip_duration)) %>% 
  ggplot(aes(mday, hour, fill = trip_duration_sum)) +
  geom_tile() +
  scale_y_reverse(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +
  facet_grid(year ~ month) +
  coord_equal() +
  scale_fill_viridis()
#need to make year strip text go horizontal
#maybe make this figure larger
```